Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-08-11 Multiresolution_framework/ltx/response_to_reviews.tex

        \item Also, Appendix A is the analogue of Appendix D of Freestone et al. and Appendix C includes results known from the literature (as the authors point out).

\emph{\parham{In order to formulate the state-space representation of the multi-resolution neural field model the inner product of the B-spline functions are required. We believe that it is necessary to provide the interested reader with the required tools to implement the algorithm. Appendix D of Freestone et al. and Appendix A herein are only similar in the sense that the inner product of two basis functions are calculated. In Appendix D of Freestone et al. two  n-dimensional Gaussians are considered while here the inner product of two B-spline functions is calculated. As for Appendix C, although authors have no objections to remove the presented derivations, we think that it would be informative for the broad audience of NeuroImage to include the derivations of some of the well-known formulations in the EM algorithm.} }

        \item Although I value the use of a wavelet as opposed to a Gaussian basis, the advantages in the context of the analysis of brain imaging data have not been demonstrated and a comparison with the authors' earlier approach has not been performed.

				\item{The framework of Freestone et al., seemed not to assume a linear relation between mean membrane potential and firing rate and in that sense the present approach appears more restrictive. }

@@

				% \emph{\parham{some notes to be added to the manuscript: Parsimony: as few as necessary variables to describe a phenomenon, complexity: to add detail description to the model to increase its accuracy with the natural phenomenon, finding a balance between complexity and parsimony is crucial. By parsimony we sacrifice realism ...  }}

				\emph{\ken{To clarify this comment, a short paragraph has been included in section 6 to highlight the modeling choice facing every user, that is; choosing between a more parsimonious or more complex model structure.}}

				\item Also, the comments on large scale networks and delays made on p.20 look irrelevant to the present framework.

				\emph{\parham{In response to this point, we have omitted the last paragraph of the discussion on delays in large scale networks.}}

### 2012-08-12 Multiresolution_framework/ltx/response_to_reviews.tex

        \item Section 5 is a review of known results and standard background material from wavelet theory and includes a simple calculation of the Fourier transform of the particular basis used here, which would be more appropriate for an appendix (cf appendix H of the Freestone et al. paper where the Fourier transform of the Gaussian basis functions is calculated).

		\emph{\parham{Section 5 summarizes important characteristics of B-Spline scaling and wavelet functions which are collected from different references, with an attempt to clarify the inner product and convolution of theses functions (please see R2 comment on ``B-spline calculus''). We think that this brief review can help the interested readers to get familiar with the B-spline scaling and wavelet functions and direct them to the important literature on this matter. As the reviewer suggested the calculation of the Fourier transform is now moved to the appendix.}}

        \item Section 6 also borrows heavily from Dewar et al, e.g. cf equation (58) of this paper with equation (19) of Dewar et al: the state transition matrix is a perturbation of the corresponding matrix used in that earlier paper. The approach used in both papers seems to be the same with an RTS smoother used in the E-step (this was also used and explained in Freestone et al.) and an M step optimizing a cost function, quadratic in the parameters, cf. equation (62) with  equation (29) of Theorem 2 of Dewar et al. In general, the derivations of section 6 and appendix B follow closely or are very similar to those of Dewar et al.

		\emph{\parham{As the reviewer correctly pointed out there are similarities in the derivations herein and that of Dewar et al. Our main aim here is to adapt techniques from control systems theory and machine learning, modify and apply them to the Amari type neural field models. As the result some of the derivations are somehow similar, however, there are important changes in the framework highlighted as follows. The inverse framework in Dewar et al. is based on a general IDE model which is not suitable when electrophysiological data is considered.  The inclusion of the membrane time constant and the firing rate in this work modify the general IDE equations to account for this type of data. As the result of these necessary modifications  the membrane parameter $\xi=1-T_s/\tau$ and the firing rate slope $\varsigma$ appear in the transition matrix. The novel implementation of the M-step in this work significantly simplifies the computational complexity of the EM based algorithm proposed by Dewar et al. as it is shown in Table.~4 of the manuscript, namely from $O(n_x^4n_{\theta}^2)$ to $O(n_x^2n_{\theta}^2)$, see equation (64) of this paper and equation (31) of Dewar et al, where the static and dynamic components are separated by defining $\boldsymbol \Gamma_1$ and $\mathbf U$ (see equations (65) and (67)). Therefore only the dynamic elements are updated at each iteration  to calculate $\boldsymbol\Upsilon$. The same simplification is also performed to calculate $\boldsymbol\upsilon_0, \boldsymbol\upsilon_1$ (see equations (62) and (63) of this paper and equation (30) of Dewar et al.). These modifications are significant and crucial for high dimensional systems and more specifically for implementing the multi-resolution framework introduced here which otherwise would be possible. The efficient M-step provided herein is also applicable in the framework of Freestone et al. if only the linear regime of the firing rate is considered.  The use of the RTS smoother in the E-step of the EM algorithm for state-space models is an effective, standard and widely used approach. In fact, the aim of the IDE representation in a state-space form is to adapt the EM and consequently the RTS smoother in the E-step. The RTS algorithm is presented only for completeness and it is different than Unscented RTS (for non-linear state-space models) in Freestone et al. The cost function of a linear state-space model with separable parameters is always quadratic in parameters in the M-step and in fact it is one of the advantages of this representation, giving a closed form solution for system's parameters which is essentially equivalent to minimum least squares error. To clarify the advantage of the M-step over the implementation in Dewar et al. we have added a paragraph to Section 8.1.}}

### 2012-08-12 Multiresolution_framework/ltx/manuscript.tex
added a paragraph to Section 8.1

One snap shot of the field reconstruction, at different levels of approximation is given in Fig.~\ref{fig:FieldEstimates100}, showing small discrepancy between the actual and estimated field at $j=3$. From this figure, it can be observed that the overall trend of the underlying field can be also captured at the coarser levels.
\subsection{Experiment II}
To study the performance of different models with various MRA capabilities, a single realization was obtained using Eqs.~\eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation} under the same circumstances of the previous experiment. The generated data set was then employed to estimate the underlying field and the connectivity kernel parameters using different state-space models, $j=0$ up to $j=4$. The mean over time of the root mean square error (MRMSE) over space of the field estimation for different models is shown in Fig.~\ref{fig:RMSE}. By increasing the value of $j$, error in the estimation decreases and converges to a steady value of 0.71, confirming that the model with $j=3$ (used in the first experiment) is in fact adequate to capture the dynamics of the underlying field. The spatiotemporal characteristics of the reconstructed fields (A-E) along with the actual field (F)  are shown in Fig.~\ref{fig:FieldEstimation}. The experiment was performed several times and the results were consistent over each run.
\section{Discussion}
In this paper, we have presented a novel model-based framework for estimating cortical dynamics from electrophysiological measurements. The novel and key developments of the paper include the multi-resolution finite element model representation of the neural field, and the estimator for an intracortical connectivity kernel with an arbitrary shape. By decomposing both the spatial field and the connectivity kernel using a wavelet decomposition, it becomes possible to represent small scale details in the field at the same time as large scale details. This work is significant, as the ability to create patient-specific neural field models has the potential to contribute to our understanding and improve treatment of diseases resulting from abnormal neurodynamics. Other groups have previously highlighted the importance of a multi-resolution approach in neural field modeling, where the dynamics and connectivity structure differs at different spatial scales~\citep{Jirsa2002,Jirsa2004,Breakspear2005,Qubbaj2009,Schultze-Kraft2011}. 
\subsection{Estimation algorithm}
For a larger space, or in a case where the connectivity kernel decomposition comprises of  many scale levels, the number of parameters to be estimated increases significantly. One solution to this problem could be expectation-conditional maximization (ECM) \citep{Meng1993,Meng1994}, which replaces the M-step by a series of computationally simplified conditional maximization (CM) steps. The ECM is a class of generalized EM (GEM)  algorithms in which the $\mathcal{Q}$-function is increased rather than being maximized \citep{Fessler1994}. For systems with a high number of states, the ensemble Kalman smoother (EnKS) \citep{Evensen2003,Evensen2009a,Evensen2009} provides an alternative to the RTS smoother used in this work. The EnKS is a sequential Monte Carlo (MC) method where the state covariance matrix is approximated by a large stochastic ensemble of model states, thus alleviating computational load due to the storage and forward integration of the state covariance matrix. However, combining the EnKS with the EM algorithm suffers from the loss of monotonicity property --- increase in the likelihood function--- as MC error will be introduced at the E-step.

>>>>

One snap shot of the field reconstruction, at different levels of approximation is given in Fig.~\ref{fig:FieldEstimates100}, showing small discrepancy between the actual and estimated field at $j=3$. From this figure, it can be observed that the overall trend of the underlying field can be also captured at the coarser levels.
\subsection{Experiment II}
To study the performance of different models with various MRA capabilities, a single realization was obtained using Eqs.~\eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation} under the same circumstances of the previous experiment. The generated data set was then employed to estimate the underlying field and the connectivity kernel parameters using different state-space models, $j=0$ up to $j=4$. The mean over time of the root mean square error (MRMSE) over space of the field estimation for different models is shown in Fig.~\ref{fig:RMSE}. By increasing the value of $j$, error in the estimation decreases and converges to a steady value of 0.71, confirming that the model with $j=3$ (used in the first experiment) is in fact adequate to capture the dynamics of the underlying field. The spatiotemporal characteristics of the reconstructed fields (A-E) along with the actual field (F)  are shown in Fig.~\ref{fig:FieldEstimation}. The experiment was performed several times and the results were consistent over each run.
\section{Discussion}
In this paper, we have presented a novel model-based framework for estimating cortical dynamics from electrophysiological measurements. The novel and key developments of the paper include the multi-resolution finite element model representation of the neural field, and the estimator for an intracortical connectivity kernel with an arbitrary shape. By decomposing both the spatial field and the connectivity kernel using a wavelet decomposition, it becomes possible to represent small scale details in the field at the same time as large scale details. This work is significant, as the ability to create patient-specific neural field models has the potential to contribute to our understanding and improve treatment of diseases resulting from abnormal neurodynamics. Other groups have previously highlighted the importance of a multi-resolution approach in neural field modeling, where the dynamics and connectivity structure differs at different spatial scales~\citep{Jirsa2002,Jirsa2004,Breakspear2005,Qubbaj2009,Schultze-Kraft2011}. 
\subsection{Estimation algorithm}
\parham{The implementation of the M-step in this work significantly simplifies the computational complexity of the EM based algorithm proposed by \cite{Dewar2009} as it is shown in Table~\ref{table:MRA-ComputationalComplexity}, namely from $O(n_x^4n_{\theta}^2)$ to $O(n_x^2n_{\theta}^2)$. This is crucial for high dimensional systems and more specifically for implementing the multi-resolution framework introduced herein which otherwise would be possible.}

For a larger space, or in a case where the connectivity kernel decomposition comprises of  many scale levels, the number of parameters to be estimated increases significantly. One solution to this problem could be expectation-conditional maximization (ECM) \citep{Meng1993,Meng1994}, which replaces the M-step by a series of computationally simplified conditional maximization (CM) steps. The ECM is a class of generalized EM (GEM)  algorithms in which the $\mathcal{Q}$-function is increased rather than being maximized \citep{Fessler1994}. For systems with a high number of states, the ensemble Kalman smoother (EnKS) \citep{Evensen2003,Evensen2009a,Evensen2009} provides an alternative to the RTS smoother used in this work. The EnKS is a sequential Monte Carlo (MC) method where the state covariance matrix is approximated by a large stochastic ensemble of model states, thus alleviating computational load due to the storage and forward integration of the state covariance matrix. However, combining the EnKS with the EM algorithm suffers from the loss of monotonicity property --- increase in the likelihood function--- as MC error will be introduced at the E-step.
